\documentclass{article}

\usepackage{amsmath,amssymb,amsthm}

\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{Week 6: Trees}
\maketitle

\begin{tabular}{c c c}
  notation & definition & recursive relationship \\
  \hline
  $t$ & binary tree \\
  $|t|$ & no. of internal nodes in tree
    & $|t| = |t_L| + |t_R| + 1$ \\
  $\boxed{t}$ & no. of external nodes in tree
    & $\boxed{t_L} + \boxed{t_R}$ \\
  $t_L$ and $t_R$ & left and right subtrees of $t$
    & \\
  $\mathrm{ipl}(t)$ & internal path length of $t$
    & $\mathrm{ipl}(t) = \mathrm{ipl}(t_L) + \mathrm{ipl}(t_R) + |t| - 1$ \\
  $\mathrm{xpl}(t)$ & external path length of $t$
    & $\mathrm{xpl}(t) = \mathrm{xpl}(t_L) + \mathrm{xpl}(t_R) + \boxed{t}$
\end{tabular}

\begin{lemma}$\boxed{t} = |t| + 1$.\end{lemma}
\begin{lemma}$\mathrm{xpl}(t) = \mathrm{ipl}(t) + 2|t|$.\end{lemma}

\section{Expected Path Length (Binary Trees)}

Let OGF $T(z) = \sum_{t \in T} z^{|t|}$ be the no. of binary trees of size
$N$ and cumulative cost OGF $Q(z) = \sum_{t \in T} \mathrm{ipl}(t)z^{|t|}$ be
the total ipl of all binary trees of size $N$. By the recursive definition of
ipl, \begin{align*}
  Q(z) &= \sum_{t_L \in T} \sum_{t_R \in T} (\mathrm{ipl}(t_L) +
  \mathrm{ipl}(t_R) + |t_L| + |t_R|)z^{|t_L| + |t_R| + 1} \\
       &= 2zT(z)(Q(z) + zT'(z)).
\end{align*} This leads to the result \begin{equation*}
  Q(z) = \frac{2z^2T(z)T'(z)}{1 - 2zT(z)}
\end{equation*} which follows to \begin{equation*}
  zQ(z) = \frac{z}{1 - 4z} - \frac{1 - z}{\sqrt{1 - 4z}} + 1.
\end{equation*} Then, $Q_N \equiv [z^N]Q(z) \sim 4^N$, and $Q_N/T_N \sim
N\sqrt{\pi N}$.

\section{Expected Path Length (Binary Search Tree)}

Let $P$ be the set of all permutations with EGF $P(z) = \sum_{p \in P}
\frac{z^{|p|}}{|p|!} = \sum_{N \geq 0} N!\frac{z^N}{N!} = \frac{1}{1 - z}$, and
cumulative cost EGF $C(z) = \sum_{p \in P} \mathrm{ipl}(p)\frac{z^{|p|}}{|p|!}$
be the total ipl built from all permutations. The expected ipl of a BST built
from a random permutation is \begin{equation*}
  \frac{N![z^N]C(N)}{[z^N]P(z)} = \frac{N![z^N]C(z)}{N!} = [z^N]C(z).
\end{equation*} A recursive decomposition of $C(z)$ yields \begin{equation*}
  C(z) = \sum_{p_L \in P} \sum_{p_R \in P} \binom{|p_L| + |p_R|}{|p_L|}
    \frac{z^{|p_L| + |p_R| + 1}}{(|p_L| + |p_R| + 1)!} (\mathrm{ipl}(p_L) +
    \mathrm{ipl}(p_R) + |p_L| + |p_R|)
\end{equation*} which when differentiated yields \begin{align*}
  C'(z) &= \sum_{p_L \in P} \sum_{p_R \in P} \frac{z^{|p_L|}}{|p_L|!}
    \frac{z^{|p_R|}}{|p_R|!}(\mathrm{ipl}(p_L) + \mathrm{ipl}(p_R) + |p_L| +
    |p_R|) \\
        &= 2C(z)P(z) + 2zP'(z)P(z) \\
        &= \frac{2C(z)}{1 - z} + \frac{2z}{(1 - z)^3}.
\end{align*} Solving for the differential equation produces \begin{equation*}
  C(z) = \frac{2}{(1 - z)^2}\ln\frac{1}{1 - z} - \frac{2z}{(1 - z)^2}
\end{equation*} and hence $C_N = 2(N + 1)(H_{N + 1} - 1) - 2N \sim 2N\ln N$.

\section*{Exercises}

\paragraph{6.6} What proportion of the forests with $N$ nodes have no trees
consisting of a single node? For $N = 1, 2, 3, 4$, the answers are
$0, 1/2, 2/5, 3/7$ respectively.

\paragraph{Solution:} Forests without singleton trees can be represented as
$SEQ(G - Z)$ where $G$ is the class of all trees. The generating function is
\begin{equation*}
  H(z) = \frac{1}{1 - (G(z) - z)} \sim \frac{4^{N + 1}}{9\sqrt{\pi N^3}},
\end{equation*} while \begin{equation*}
  G(z) = \frac{1 - \sqrt{1 - 4z}}{2} \sim \frac{4^N}{\sqrt{\pi N^3}},
\end{equation*} so $H(z)/G(z) \sim 4/9$.

\paragraph{6.27} When $N$ is a power of two minus 1, what is the probability
that a perfectly balanced tree structure (all external nodes on the same level)
will be built, if all key insertion sequences are equally likely?

\paragraph{Solution:} Let $a_n$ be the number of permutations corresponding to
a perfectly balanced BST with depth $n$, and $b_n$ be the number of
permutations of size $N = 2^n - 1$. Then, \begin{gather*}
  a_n = \binom{2(2^{n - 1} - 1)}{2^{n - 1} - 1}a_{n - 1}^2, \\
  b_n = (2^n - 1)! = (2^n - 1)\binom{2(2^{n - 1} - 1)}{2^{n - 1} - 1}b_{n-1}^2,
\end{gather*} which leads to \begin{equation*}
  \frac{a_n}{b_n} = \frac{(a_{n - 1}/b_{n - 1})^2}{2^n - 1} =
  \prod_{j = 0}^{n - 1} \frac{1}{(2^{n - j} - 1)^{2^j}}
\end{equation*} as $a_1 = b_1 = 1$.

\paragraph{6.43} Internal nodes in binary trees fall into one of three classes:
they have either two, one, or zero external children. What fraction of the
nodes are of each type, in a random binary search tree of $N$ nodes?

\paragraph{Solution:} Let $l(\sigma)$ be the number of leaves (internal nodes
with zero external children) of the BST corresponding to permutation $\sigma$.
The cumulative EGF is \begin{align*}
  C(z) &= \sum_{\sigma} \frac{l(\sigma)}{|\sigma|!}z^{|\sigma|} \\
       &= z + \sum_{\sigma_L} \sum_{\sigma_R}
       \frac{1}{(|\sigma_L| + |\sigma_R| + 1)!}
       \binom{|\sigma_L| + |\sigma_R|}{|\sigma_L|}(l(\sigma_L) + l(\sigma_R))
       z^{|\sigma_L| + |\sigma_R| + 1}.
\end{align*} Taking the derivative on both sides and simplifying, we have
\begin{equation*}
  C'(z) = 1 + \frac{2}{1 - z}C(z)
\end{equation*} which solves to \begin{equation*}
  C(z) = \frac{1}{3}\left(\frac{1}{(1 - z)^2} - (1 - z)\right),
\end{equation*} yielding $C_N = \frac{1}{3}(n + 1)!$ for $n \geq 2$, so
the average number of leaves is asymptotically $\frac{n + 1}{3}$.

Similarly, let $r(\sigma)$ be the number of internal nodes with two external
children of the BST corresponding to permutation $\sigma$. The cumulative EGF
is \begin{align*}
  D(z) &= \sum_{\sigma} \frac{r(\sigma)}{|\sigma|!}z^{|\sigma|} \\
       &= \sum_{|\sigma_L| > 0} \sum_{|\sigma_R| > 0}
       \frac{1}{(|\sigma_L| + |\sigma_R| + 1)!}
       \binom{|\sigma_L| + |\sigma_R|}{|\sigma_L|}(r(\sigma_L) + r(\sigma_R) +
       1)z^{|\sigma_L| + |\sigma_R| + 1} \\
       &+ 2\sum_{\sigma} \frac{r(\sigma)}{(|\sigma| + 1)!}z^{|\sigma| + 1}
\end{align*} which simplifies considerably to \begin{align*}
  D(z) &= \sum_{\sigma_L} \sum_{\sigma_R}
       \frac{1}{(|\sigma_L| + |\sigma_R| + 1)!}
       \binom{|\sigma_L| + |\sigma_R|}{|\sigma_L|}(r(\sigma_L) + r(\sigma_R) +
       1)z^{|\sigma_L| + |\sigma_R| + 1} \\
       &-2\sum_{\sigma} \frac{1}{(|\sigma| + 1)!}z^{|\sigma| + 1} + z.
\end{align*} Taking derivatives and simplifying, we have \begin{equation*}
  D'(z) = \frac{2}{1 - z}D(z) + \frac{z^2}{(1 - z)^2}
\end{equation*} which solves to \begin{equation*}
  D(z) = \frac{z^3}{3(1 - z)^2}
\end{equation*} yielding $D_N = \frac{1}{3}(n - 2)n!$ for $n \geq 2$, so the
average number of leaves is asymptotically $\frac{n - 2}{3}$.

Subtracting away these two numbers, the average number of internal nodes with
one external child is $\frac{n + 1}{3}$ for $n \geq 2$.

\end{document}

