\documentclass{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage[shortlabels]{enumitem}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}

\title{Chapter 2: Numbers of Various Sorts}
\maketitle

\paragraph{Mathematical Induction.} That's the whole chapter, it's just about
natural numbers and mathematical induction.

\section*{Exercises}

\paragraph{Problem 7} The base case $p = 0$ is trivial. To construct an
inductive hypothesis, consider the formula
\begin{multline*}
  (k + 1)^{p+1} - k^{p+1} \\
  = \binom{p+1}{1} \cdot k^p + \binom{p+1}{2} \cdot k^{p-1} + \cdots +
  \binom{p+1}{p-1} \cdot k^2 + \binom{p+1}{p} \cdot k + 1.
\end{multline*}
Writing this formula for $k = 1, \ldots, n$ yields
\begin{multline*}
  2^{p+1} - 1^{p+1} \\
  = \binom{p+1}{1} \cdot 1^p + \binom{p+1}{2} \cdot 1^{p-1} + \cdots +
  \binom{p+1}{p-1} \cdot 1^2 + \binom{p+1}{p} \cdot 1 + 1,
\end{multline*}
\begin{multline*}
  3^{p+1} - 2^{p+1} \\
  = \binom{p+1}{1} \cdot 2^p + \binom{p+1}{2} \cdot 2^{p-1} + \cdots +
  \binom{p+1}{p-2} \cdot 2^2 + \binom{p+1}{p-1} \cdot 2 + 1,
\end{multline*}
\[
  \vdots
\]
\begin{multline*}
  (n + 1)^{p+1} - n^{p+1} \\
  = \binom{p+1}{1} \cdot n^p + \binom{p+1}{2} \cdot n^{p-1} + \cdots +
  \binom{p+1}{p-2} \cdot n^2 + \binom{p+1}{p-1} \cdot n + 1.
\end{multline*}
Adding together, we obtain
\begin{multline*}
  (n + 1)^{p+1} - 1^{p+1} \\
  = \binom{p+1}{1} \cdot (1^p + 2^p + \cdots + n^p) + \binom{p+1}{2} \cdot
  (1^{p-1} + 2^{p-1} + \cdots + n^{p-1}) \\ + \cdots + \binom{p+1}{p-1} \cdot
  (1^2 + 2^2 + \cdots + n^2) + \binom{p+1}{p} \cdot (1 + 2 + \cdots + n) + n
\end{multline*}
which can be rewritten as
\begin{multline*}
  (n+1)^{p+1} - 1^{p+1} \\
  = \binom{p+1}{1} \sum^n_{k=1} k^p + \binom{p+1}{2} \sum^n_{k=1} k^{p-1} +
  \cdots \\
  + \binom{p+1}{p-1} \sum^n_{k=1} k^2 + \binom{p+1}{p} \sum^n_{k=1} k + n.
\end{multline*}
It then follows naturally that, by expansion of $(n + 1)^{p+1}$ and assuming
that the statement to be proven holds for all sums of lower order than $p$,
that \[
  n^{p+1} + f(n) = \binom{p+1}{1} \sum^n_{k=1} k^p
\] where $f(n)$ is a polynomial of degree $p$ in $n$. Clearly, the statement is
proven true.

\paragraph{Problem 17}
\begin{enumerate}[(a)]
  \item Suppose that every number at least 2 and less than $n$ can be written
    as a product of primes. Clearly, if $n$ is a prime, it can be written as a
    product of primes (namely, itself). If $n$ is not a prime, then $n = ab$
    for some natural numbers $2 \leq a, b < n$. By the assumption that $a$ and
    $b$ can be written as products of primes, $n = ab$ can also be written as a
    product of primes.
  \item For some rational $\sqrt{n}$ i.e. $\sqrt{n} = a/b$ for some natural
   numbers $a, b$, then $b\sqrt{n} = a$ and their prime factorisations must be
   the same. Squaring both sides, $b^2n = a^2$. Since every prime appears an
   even number of times in the factorisation of $a^2$, and of $b^2$, the same
   must be true of the factorisation of $n$. Hence, $n$ must be a square. This
   proves the statement.
  \item Applying the same argument in (b), for some rational $\sqrt[k]{n}$ i.e.
    $\sqrt[k]{n} = a/b$ for some natural numbers $a, b$, $b^kn = a^k$. Since
    every prime appears a multiple of $k$ times in the factorisation of $a^k$,
    and of $b^k$, the same must be true of the factorisation of $n$. Hence, $n
    = m^k$ for some natural number $m$. This proves the statement.
  \item Suppose that there are a finite number of primes, as stated in (d).
    Then, $p_1 \cdot p_2 \cdot \ldots \cdot p_n + 1$ would divide each prime
    $p_1, p_2, p_3, \ldots, p_n$ with remainder 1. Should this be a prime, then
    there is a contradiction; should this be a composite number, by the unique
    factorisation theorem, the number can be rewritten into a product of primes
    none of which are $p_1, p_2, p_3, \ldots, p_n$, leading into a
    contradiction. Hence, there must be an infinite number of primes.
\end{enumerate}

\paragraph{Problem 18.}
\begin{enumerate}[(a)]
  \item For some rational solution $x$ i.e. $x = b/c$ for some integers $b, c$
    with no common factors, \[
      \frac{b^n}{c^n} + a_{n-1} \cdot \frac{b^{n-1}}{c^{n-1}} + \cdots + a_0
      \cdot \frac{b^0}{c^0} = 0
    \] which can be multiplied by $c^n$ and rewritten to yield \[
      b^n + a_{n-1} \cdot b^{n-1}c^1 + \cdots + a_0 \cdot c^n = 0.
    \] If $c \neq \pm 1$, then $c$ has some prime number as a factor. This
    prime factor divides every term other than $b^n$, so it must divide $b^n$
    too. However, since $b$ and $c$ share no common factors, the prime factor
    cannot divide $b^n$. Hence, $c = \pm 1$. This proves that $x$ is an
    integer.
  \item Let $x = \sqrt{6} - \sqrt{2} - \sqrt{3}$. Then,
    \begin{align*}
      x^2 &= 6 + (\sqrt{2} + \sqrt{3})^2 - 2\sqrt{6}(\sqrt{2} + \sqrt{3}) \\
          &= 11 + 2\sqrt{6}[1 - (\sqrt{2} + \sqrt{3})],
    \end{align*}
    which implies
    \begin{align*}
      (x^2 - 11)^2
      &= 24[1 - (\sqrt{2} + \sqrt{3})]^2 \\
      &= 24[1 + (\sqrt{2} + \sqrt{3})^2 - 2(\sqrt{2} + \sqrt{3})] \\
      &= 24[6 + 2\sqrt{2}\sqrt{3} - 2(\sqrt{2} + \sqrt{3})] \\
      &= 24(6 + 2x).
    \end{align*}
    From (a), $x$ is irrational unless it is an integer. It is easy to check
    that $0 < \sqrt{6} - \sqrt{2} - \sqrt{3} < 1$, and so $x$ must be
    irrational.
  \item Let $x = \sqrt{2} + \sqrt[3]{2} = \gamma^2 + \gamma^3$ where $\gamma =
    2^{1/6}$. Considering the first sixth powers of $x$ and constructing a
    table for the coefficients of each term,

    \begin{tabular}{c | c c c c c c}
      & $\gamma^0$ & $\gamma^1$ & $\gamma^2$ & $\gamma^3$ & $\gamma^4$ &
      $\gamma^5$ \\
      \hline
      $x^0$ & 1 & & & & & \\
      $x^1$ & & & 1 & 1 & & \\
      $x^2$ & 2 & & & & 1 & 2 \\
      $x^3$ & 2 & 6 & 6 & 2 & & \\
      $x^4$ & 4 & & 2 & 8 & 12 & 8 \\
      $x^5$ & 40 & 40 & 20 & 4 & 2 & 10 \\
      $x^6$ & 12 & 24 & 60 & 80 & 60 & 24 \\
    \end{tabular}

    We can then find numbers $a_0, \ldots, a_5$ such that \[
      x^6 + a_5x^5 + \cdots + a_0 = 0,
    \] by substituting in $x^0, x^1, \ldots, x^6$ in terms of $\gamma$. This
    leaves us with the following matrix: \[
      \begin{bmatrix}
        1 & 0 & 2 & 2 & 4 & 40 & 12 \\
        0 & 0 & 0 & 6 & 0 & 40 & 24 \\
        0 & 1 & 0 & 6 & 2 & 20 & 60 \\
        0 & 1 & 0 & 2 & 8 & 4 & 80 \\
        0 & 0 & 1 & 0 & 12 & 2 & 60 \\
        0 & 0 & 2 & 0 & 8 & 10 & 24 \\
      \end{bmatrix}
      \begin{bmatrix}
        a_0 \\
        a_1 \\
        a_2 \\
        a_3 \\
        a_4 \\
        a_5 \\
        1 \\
      \end{bmatrix}
      = 0
      \] which turns out to have a solution corresponding to the equation \[
        x^6 + 6x^5 + 4x^4 - 12x^3 + 24x^2 + 4x = 0.
      \] Part (a) implies that either $x$ is irrational or $x$ is an integer,
      and it is easy to see that $x$ is not an integer, so $x$ must be
      irrational.
\end{enumerate}

\end{document}

