\documentclass{article}

\usepackage{amsmath,amssymb,amsthm}

\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem*{lemma*}{Lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{theorem*}{Theorem}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{Chapter 16: $\pi$ is Irrational}
\maketitle

Before the proof, two lemmas are required.

\begin{lemma*}
  The function \begin{equation*}
    f_n(x) = \frac{x^n(1 - x)^n}{n!}
  \end{equation*} satisfying \begin{equation*}
    0 < f_n(x) < \frac{1}{n!} \text{ for } 0 < x < 1
  \end{equation*} satisfies various properties: \begin{enumerate}
    \item $f_n^{(k)}(0) = f_n^{(k)}(1) = 0$ if $k < n$ or $k > 2n$, and
    \item $f_n^{(k)}(0)$ and $f_n^{(k)}(1)$ are integers for all $k$.
  \end{enumerate}
\end{lemma*}

\begin{proof}
  We first write $f_n$ in the form \begin{equation*}
    f_n(x) = \frac{1}{n!}\sum_{i = n}^{2n} c_ix^i
  \end{equation*} for integers $c_1, c_2, \ldots, c_n$. Clearly,
  \begin{equation*}
    f_n^{(k)}(0) = 0 \text{ if } k < n \text{ or } k > 2n.
  \end{equation*} Moreover, \begin{align*}
    f_n^{(n)}(x) &= \frac{1}{n!}[n!c_n + \text{ terms involving } x] \\
    f_n^{(n + 1)}(x) &= \frac{1}{n!}[(n + 1)!c_{n + 1} +
    \text{ terms involving } x] \\
    \vdots& \\
    f_n^{(2n)}(x) &= \frac{1}{n!}[(2n)!c_{2n}].
  \end{align*} which implies that \begin{align*}
    f_n^{(n)}(0) &= c_n \\
    f_n^{(n + 1)}(0) &= (n + 1)c_{n + 1} \\
    \vdots& \\
    f_n^{(2n)}(0) &= (2n)(2n - 1)\cdots(n + 1)c_{2n},
  \end{align*} where the numbers on the right are all integers. Thus
  $f_n^{(k)}(0)$ are integers for all $k$.

  Furthermore, the relation $f_n(x) = f_n(1 - x)$ implies that $f_n^{(k)}(x) =
  (-1)^kf_n^{(k)}(1 - x)$; therefore, $f_n^{(k)}(1) = 0$ if $k < n$ or $k >
  2n$, and is also an integer for all $k$.
\end{proof}

\begin{lemma*}
  If $a$ is any number, and $\epsilon > 0$, then for sufficiently large $n$ we
  have \begin{equation*}
    \frac{a^n}{n!} < \epsilon.
  \end{equation*}
\end{lemma*}

\begin{proof}
  Notice that if $n \geq 2a$, then \begin{equation*}
    \frac{a^{n + 1}}{(n + 1)!} = \frac{a}{n + 1} \cdot \frac{a^n}{n!} <
    \frac{1}{2} \cdot \frac{a^n}{n!}.
  \end{equation*} Now let $n_0$ be any natural number with $n_0 \geq 2a$. Then,
  whatever value \begin{equation*}
    \frac{a^{n_0}}{(n_0)!}
    \end{equation*} may have, the succeeding values satisfy \begin{align*}
    \frac{a^{n_0 + 1}}{(n_0 + 1)!} &< \frac{1}{2} \cdot \frac{a^{n_0}}{n_0!} \\
    \frac{a^{n_0 + 2}}{(n_0 + 2)!} &< \frac{1}{2} \cdot
                                   \frac{a^{n_0 + 1}}{(n_0 + 1)!}
                                   < \frac{1}{2} \cdot \frac{1}{2} \cdot
                                   \frac{a^{n_0}}{n_0!} \\
    \vdots& \\
    \frac{a^{n_0 + k}}{(n_0 + k)!} &< \frac{1}{2^k} \cdot \frac{a^{n_0}}{n_0!}.
  \end{align*} If $k$ is so large that $\frac{a^{n_0}}{(n_0)!\epsilon} < 2^k$,
  then \begin{equation*}
    \frac{a^{n_0 + k}}{(n_0 + k)!} < \epsilon.
  \end{equation*}
\end{proof}

\begin{theorem}
  The number $\pi$ is irrational; in fact, $\pi^2$ is irrational. (Notice that
  the irrationality of $\pi^2$ implies the irrationality of $\pi$, for if $\pi$
  were rational, then $\pi^2$ certainly would be.)
\end{theorem}

\begin{proof}
  Suppose $\pi^2$ were rational, so that \begin{equation*}
    \pi^2 = \frac{a}{b}
    \end{equation*} for some positive integers $a$ and $b$. Let
  \begin{equation}
    G(x) = b^n[\pi^{2n}f_n(x) - \pi^{2n - 2}f_n''(x) - \cdots +
    (-1)^nf_n^{(2n)}(x)].
  \end{equation} Notice that each of the factors \begin{equation*}
    b^n\pi^{2n - 2k} = b^n(\pi^2)^{n - k} =
    b^n\left(\frac{a}{b}\right)^{n - k} = a^{n - k}b^k
  \end{equation*} is an integer. Since $f_n^{(k)}(0)$ and $f_n^{(k)}(1)$ are
  integers, this shows that $G(0)$ and $G(1)$ are integers. Differentiating
  $G$ twice yields \begin{equation}
    G''(x) = b^n[\pi^{2n}f_n''(x) - \pi^{2n - 2}f_n^{(4)}(x) - \cdots +
    (-1)^nf_n^{(2n + 2)}(x)].
  \end{equation} The last term, $(-1)^nf_n^{(2n + 2)}(x)$, is zero. Thus,
  adding (1) and (2) gives \begin{equation}
    G''(x) + \pi^2G(x) = b^n\pi^{2n + 2}f_n(x) = \pi^2a^nf_n(x).
  \end{equation} Now let \begin{equation*}
    H(x) = G'(x)\sin\pi x - \pi G(x) \cos\pi x.
  \end{equation*} Then \begin{align*}
    H'(x) &= \pi G'(x)\cos\pi x + G''(x)\sin\pi x - \pi G'(x) \cos\pi x + \pi^2
    G(x)\sin\pi x \\
          &= [G''(x) + \pi^2 G(x)]\sin\pi x \\
          &= \pi^2a^nf_n(x)\sin\pi x, \text{ by (3)}.
  \end{align*} By the Second Fundamental Theorem of Calculus, \begin{align*}
    \pi^2\int_0^1 a^nf_n(x)\sin\pi x \,\mathrm{d}x
    &= H(1) - H(0) \\
    &= G'(1)\sin\pi - \pi G(1)\cos\pi - G'(0)\sin 0 + \pi G(0)\cos 0 \\
    &= \pi[G(1) + G(0)].
  \end{align*} Thus \begin{equation*}
    \pi\int_0^1 a^nf_n(x)\sin\pi x \,\mathrm{d}x \text{ is an integer}.
  \end{equation*} On the other hand, $0 < f_n(x) < 1/n!$ for $0 < x < 1$, so
  \begin{equation*}
    0 < \pi a^nf_n(x)\sin\pi x < \frac{\pi a^n}{n!} \text{ for } 0 < x < 1.
  \end{equation*} Consequently, \begin{equation*}
    0 < \pi \int_0^1 a^nf_n(x)\sin\pi x \,\mathrm{d}x < \frac{\pi a^n}{n!}.
  \end{equation*} This reasoning was completely independent of the value of
  $n$. Now if $n$ is large enough, then \begin{equation*}
    0 < \pi \int_0^1 a^nf_n(x)\sin\pi x \,\mathrm{d}x < \frac{\pi a^n}{n!} < 1.
  \end{equation*} But this is absurd, because the integral is an integer, and
  there is no integer between 0 and 1. Thus our original assumption must have
  been incorrect: $\pi^2$ is irrational.
\end{proof}

\end{document}

